---
title: "An Introduction to lr2cluster"
author: "Kurnia Susvitasari"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_md: true
---

This document provides an introductory tutorial on the `lr2cluster` package. This package implements tools to help identify cluster assignments using multiple data sources when direct methods such as contact tracing and genomic sequencing are only available for some data. There are also complimentary functions, that can be used to estimate a cluster’s true size and to choose which new cases to be sequenced next. This tutorial provides an overview of `lr2cluster`’s basic functionalities.



# Installing the Package

To install the package, you need to install the package remotes first:

```{r install remote, message=FALSE, eval=FALSE}
install.packages("remotes")
```

Once remotes is successfully installed, run the following to install and load `lr2cluster`:

```{r install lr2cluster, message=FALSE, eval=F}
remotes::install_github("ksusvita92/lr2cluster")
```

```{r load lr2cluster, message=FALSE}
library(lr2cluster)
```



# Valencia Data

Data for TB cases in Valencia is available at https://doi.org/10.7554/eLife.76605. The following is a short script to prepare the data. In this tutorial, we do not provide the cases' true geographical information. The location data used in this tutorial is a toy example which can be downloaded from GitHub repository [here](https://raw.githubusercontent.com/ksusvita92/Genomic-Clustering/master/analysis%20scripts/location_data.csv).

```{r load data, message=FALSE}
library(readxl)
library(dplyr)

url <- paste("https://elifesciences.org/download/aHR0cHM6Ly9jZG4uZWxpZmVzY2llbmNlcy5vcmc",
"vYXJ0aWNsZXMvNzY2MDUvZWxpZmUtNzY2MDUtc3VwcDEtdjEueGxzeA--/elife-76605-supp1-v1.xlsx?_ha",
"sh=pzQwKD1DzDLre7kKrWI%2Fhd%2BjY2FGgpekrPI4vXrlWNo%3D", sep = "")
destfile <- "rawdt.xlsx"
curl::curl_download(url, destfile)
rawdt <- read_excel(destfile, range = "A3:AB778", na = "NA")

# subset the necessary column and cases
tb_valencia <- rawdt %>% 
  transmute(ID = ...1, Cluster = `Genomic
Cluster ID`, 
  Gender = Gender, 
  Foreign = ifelse(`Country of birth`=="SPAIN", "No", "YES"), 
  Diabetes = Diabetes, 
  HIV = `HIV infected`) %>%
  filter(!is.na(Gender), !is.na(Foreign), !is.na(Diabetes), !is.na(HIV)) 

# subset clusters having >2 members
nm <- as.data.frame(table(tb_valencia$Cluster)) %>% filter(Freq > 2) %>% pull(Var1)
tb_valencia <- tb_valencia %>% filter(Cluster %in% nm)

# download location data from a repo
url <- paste("https://raw.githubusercontent.com/ksusvita92/Genomic-Clustering/master/",
  "analysis%20scripts/location_data.csv", sep = "")
location_data <- read.csv(url)
tb_valencia <- tb_valencia %>%
  inner_join(location_data, by = "ID")
str(tb_valencia)
```




## Transformation to pairwise data

Our pairwise logistic regression model uses pairwise data, as discussed in the paper. For each variable in the raw individual data, each pair of cases has a pairwise variable that represents the dissimilarity between the two cases' individual variables. For example, each case has latitude and longitude in the individual data, but each pair of cases has distance as a variable in the pairwise data.

To transform Valencia data into pairwise data, run:

```{r transform to pairwise}
dt <- tb_valencia[tb_valencia$Cluster != "unique",] #exclude unclustered cases
pairdt <- zCovariate(cluster = dt$Cluster,
                     X = dt[,3:6],
                     location = dt[,7:8],
                     id = dt$ID)
head(pairdt)
```


Notice that there are 4 new columns in the data frame above: `case`, `to.case`, `y`, and `Spatial`.

  + `case`, `to.case`: vector of case id; together represent a pair.
  + `y`: a binary variable; `y = 1` means if a pair is in the same cluster, `y = 0` means otherwise.
  + `Spatial`: numeric vector that represents spatial distance between a pair. 
  
Each categorical variable will have one additional level, called `DIFF` which indicates if a pair has different values. To see more details about the function above, run `?zCovariate`.



# Pairwise Logistic Regression

There are two logistic regression models introduced in this package: the multinomial logistic regression (MLR) and the pairwise logistic regression (PLR). The difference between the two models lies on the data type used to train the model, and the response variable. MLR uses individual-level data and the response variable is the index of the cases’ clusters, whereas PLR uses pairwise-level data and binary response variable indicating whether two cases are in the same cluster.


## Fit PLR model

PLR perform regression analysis which maps pairwise predictors to a binary response variable. Suppose we want to fit predictors `sex`, `foreign`, `dx_data`, and `latitude` and `longitude.`

```{r fit plt}
fit_plr <- plr(formula = Cluster ~ Gender+Foreign+Diabetes+HIV+Latitude+Longitude,
               data = dt)
summary(fit_plr)
```

`summary()` function provides the estimated regression coefficients, together with the Wald test. An important note for the users is if ones want to fit spatial distance between two cases, ones must name the geographical location as `latitude` (or `lat`, `Lat`, `Latitude`), and `longitude` (or `long`, `Long`, `Longitude`).


## Predict if two cases are in the same cluster

Suppose there are a collection of new cases for which we have no information on their true clusters, and we want to predict the probability of a pair belonging to the same cluster. To illustrate this, let us split the `tb_valencia` data into two: one is to train the PLR model and the other is to test the model. 

To make sure we split the data such that there is at least one case per cluster in each data split, use `createDataPartition()` from `caret` package. In this example, we set 60% proportion of the data to be in the training set.

```{r split the data}
library(caret)

set.seed(12345)
id <- createDataPartition(dt$Cluster, p = .6, list = F)
traindt <- dt[id,]
testdt <- dt[-id,]
```

Use `plr()` function to fit PLR model on `traindt`, and run the following to get the estimated probability of the response variable:

```{r train the model and predict}
fit_plr <- plr(formula = Cluster ~ Gender+Foreign+Diabetes+HIV+Latitude+Longitude,
               data = traindt)
pred_plr <- predict(obj = fit_plr,
                    newdata = testdt,
                    case.id = testdt$ID) #case.id can be NULL
```

The function `predict.plr()` (or `predict()`) returns a data frame which contains a vector of the predicted probability of a pair of cases are in the same cluster, and its standard error.


## Finding the optimum threshold
As in any binary classification problem where a threshold must be selected, `lr2cluster` provides a function to find an optimum threshold which depends on how the user values the cost of false positive errors (saying two cases are in the same cluster when they are not) versus false negative errors (saying two cases are not in the same cluster when they are).

The function `optThreshold()` requires a vector of the true response variable and its prediction. The later is obtained from `pred_plr$y`, and to obtain the former, use `zCovariate()` function on `testdt`.

```{r opt threshold}
# get the true response variable of testdt data
tr_resp <- zCovariate(cluster = testdt$Cluster)$y
opt_threshold <- optThreshold(response = tr_resp,
                              prediction = pred_plr$y,
                              cost.ratio = 30) #default is 1
opt_threshold
```

`optThreshold()` returns a list of values, such as:

  + `threshold`: the optimum threshold obtained, given `cost.ratio`.
  + `specificity`, `sensitivity`: the true negative rate and the true positive rate, evaluated at the optimum threshold.
  + `accuracy`: accuracy of the prediction.
  + `roc``: an object of class "roc".
  
To plot the ROC, you can run:

```{r plot roc, fig.height=3, fig.width=4}
plot(opt_threshold)
```

which also shows the position of the optimum threshold given `cost.ratio`.



# Cluster Assignment

The previous section shows how to predict if two new cases belong to the same cluster. However, one may wonder if those new cases belong to one of the known clusters in the training set. This section shows how to find the most probable clusters a new case can be assigned to based on some scores called _cluster scores_. These scores, which are on a scale between 0 to 1, represent how likely a new case belong to a given cluster.



## Finding probable clusters a case most likely to belong to

For a given new case, cluster assignment can be done by setting a threshold, and selecting all clusters for which the cluster score exceeds the threshold. Alternatively, one can choose the  _K_ clusters with the highest cluster score for the case. The following code does the first if a threshold if provided, and otherwise assigns a case to the best _K_ clusters.

To do this task, we will use `traindt` and `testdt` again, so that we can test the accuracy of our assignments. We will use the same variables as the previous section as well.

```{r cluster assignment}
assgn_plr <- clusterPLR(formula = Cluster ~ Gender+Foreign+Diabetes+HIV+Latitude+Longitude,
                        data = traindt,
                        newdata = testdt,
                        threshold = NULL,
                        nbest = 3)
assgn_plr
```

The code above is to find the best _K = 3_ clusters a case can be assigned to using PLR model. If ones want to compare the method using random assignment or MLR model, run `clusterRandom()` or `clusterMLR()`.


To obtain the most probable clusters for all new cases, run:

```{r get best clusters, eval=FALSE}
getbest(assgn_plr) #or
assgn_plr$best_cluster
```

To obtain the cluster scores, run:

```{r get best cases, eval=FALSE}
getclusterScore(assgn_plr) #or
assgn_plr$cluster_score
```

See `?clusterPLR`, `?clusterRandom`, and `?clusterMLR` for more details.



## Compute the accuracy of assigning cases to their correct clusters

The accuracy in this context is the fraction of new cases whose true clusters is in the best _K_ clusters predicted. For example, suppose that a case’s true cluster is "Cluster A". If "Cluster A" is predicted as one of the _K_ clusters, then the method will consider the assignment correct.

To compute the accuracy for predicting the new cases assignment, run

```{r acc, eval=FALSE}
acc(obj = assgn_plr, true.cluster = testdt$Cluster)
```




# Other Applications

## Cases to be sequenced next

Instead of predicting clusters a case can be assigned to, we can turn this relationship around, and ask, for a given cluster, which unassigned cases are most likely to belong to it. This could be used, for example, to decide which of several unsequenced cases should be sequenced next, if we are interested in identifying all the cases that belong to a particular cluster.

To do this task, run

```{r get next cases}
next_case <- case2sequence(obj = assgn_plr,
                           case.id = testdt$ID,
                           nbest = 3)
next_case
```


The above code returns the suggested 3 cases to be chosen given a cluster based on their cluster score’s rank. If one wants to find the _K_ best cases, change _nbest = K_ or, one can also provide a threshold which serves as a cut-off to any cases with lower score.

To obtain the best cases for all clusters, run

```{r best cases, eval=FALSE}
getbest(next_case) #or
next_case$best_cases
```


To obtain the accuracy on this task, run

```{r acc cases, eval=FALSE}
acc(obj = next_case, true.cluster = testdt$Cluster)
```



## Estimate a cluster’s true size

Suppose that we have a cluster of interest _C_ with some cases in it, and a collection of unassigned new cases. We can estimate the total number of new cases that would get assigned to _C_, and therefore estimate cluster _C_’s true size.

```{r cluster size, eval=FALSE}
clusterSize(obj = assgn_plr, rho = 0)
```


The argument `rho` in `clusterSize()` represents the probability that a case does not belong to any of any given clusters.


# Closing Remarks

`lr2cluster` is a tool to assign newly identified cases of an infectious disease to existing transmission clusters using several data streams. The application is extended to also, for example, predict which new cases to be sequenced next, given a cluster, and estimate a cluster’s true size.

For general questions and bug reports, please send a message to ksusvita@gmail.com.
